\ssection{Implementation}
The implmentations created as part of this bachorlor thesis aimed
to make use of the LLVM compiler infastructure.
LLVM is a collection of modular and reusable compiler and toolchain technologies,
most notably for this project is the clang compiler and
lldb debugger. Furthermore, qemu will be used extensively while testing the
implementations.

\subsection{Dependencies}
\subsubsection{QEMU}
\begin{lstlisting}[caption=Installing QEMU, float=*, label=lst:qemu_install]
git clone git clone https://github.com/qemu/qemu  # Clone the qemu repo
./configure --target-list=riscv32-softmmu  # Configure the 32-bit RISC-V target
make -j $(nproc)  # build the project with all num cores jobs
sudo make install
\end{lstlisting}.
QEMU is a system emulator, which has the capabilities of emulating both a 32-bit
and 64-bit RISC-V CPU\cite{QEMU}.
Following the instructions by RISC-V's getting started guide
we can build the QEMU RISC-V system emulators by running the code
provided in Listing~\ref{lst:qemu_install}\cite{RISC-V_GS}.\footnote{Once
installed make sure to add both llvm and riscv gnu toolchain
to path. Both should be installed in the /opt/ folder.}

\subsubsection{Installing LLVM compiler infastructure}
\begin{lstlisting}[caption=Installing LLVM compiler infastructure with RISC-V
32-bit as native target., float=*, label=lst:llvm_install]
# Dependencies
sudo apt-get -y install \
  binutils build-essential libtool texinfo \
  gzip zip unzip patchutils curl git \
  make cmake ninja-build automake bison flex gperf \
  grep sed gawk python bc \
  zlib1g-dev libexpat1-dev libmpc-dev \
  libglib2.0-dev libfdt-dev libpixman-1-dev

# Installing the RISC-V-gnu-toolchain
git clone https://github.com/riscv-collab/riscv-gnu-toolchain  # clone
riscv-gnu-toolchain
cd riscv-gnu-toolchain  # change directory
./configure --prefix=/opt/riscv --enable-multilib
# prefix is install path used by llvm
# --enable-multilib allows us to compile for 32-bit
sudo make -j$(nproc)
cd ..

# Installing LLVM
git clone https://github.com/llvm/llvm-project.git # clone llvm-project
cd llvm-project
mkdir build
pushd build
sudo cmake -S ../llvm -G Ninja \
-DCMAKE_BUILD_TYPE="Release" -DBUILD_SHARED_LIBS=True \
-DLLVM_BUILD_TESTS=False \
-DLLVM_ENABLE_PROJECTS="clang;clang-tools-extra;lld" \
-DCMAKE_INSTALL_PREFIX=/opt/llvm \
-DLLVM_TARGETS_TO_BUILD="RISCV"
sudo cmake --build . --target install
popd
\end{lstlisting}
When it comes to clang there are two methods of installing, that are relevant to
this project. If running on a debian based system, then you can simply install
llvm-tools package. The issue with this approach is that the general build is
for use with the current system installation is on, which unless you are running
a RISC-V computer architecture natively will lead to issues when trying to cross
compile if the given targets use any of the standard libraries, such as
freeRTOS. A fix to this issue is to explecitely tell clang to make use of the
RISC-V gnu toolchain on every compilation\dots

The second approach is to build llvm with the RISC-V 32-bit target as the native
target. This approach is documented in Listing~\ref{lst:llvm_install}. After
installation it is important to add both clang build and RISC-V gnu toolchain to
PATH. However, adding the following flags to compilation
should lead to the same results, although the second approach is used throughout
this project.
\begin{itemize}
  \item --sys-root={Path to RISC-V install}/riscv64-unknown-elf
  \item --taget=riscv32
  \item --gcc-toolchain={Path to RISC-V install}
\end{itemize}


\subsection{Bare metal C application}
\subsubsection{Getting system information}\label{sec:sys-info}
Now that there is a working toolchain, we can move on to the developement of the
bare metal C version. First, we need information about the delvelopemnt
environment we are currently working on, such that we are able to setup a stack
for the bare metal C program. With QEMU it is possible to get the necesarry
machine info by running:
\begin{lstlisting}
qemu-system-riscv32 -machine virt \
-machine dumpdtb=riscv32.dtb
\end{lstlisting}
This creates a Devicetree Blob(dtb) datafile, which contains information about
the virt qemu-system-riscv32 virtual machine. This format is not usuable by us
at the moment, but by using the Device Tree Compiler(dtc) package we can convert
it from the binary dtb format to a human readable dts format.
\begin{lstlisting}
sudo apt install dtc
dtc -I dtb -O dts -o riscv32.dts riscv32.dtb
\end{lstlisting}
Opening the file up in your favorite text editor you should see a lot of
information regarding the qemu-system-riscv32 virtual machine. First we note,
that the Devicetree specification states, that the memory node descirbes the
physical memory layout for the system. As we want the programs stack
to live within the memory section, this is section we should find information
about starting address and length of the memory section. The memory node has
two required sections, first the device\_type, which must simply be "memory", and
secondly the reg value. The reg value "Consists of an arbitrary number of
address and size pairs that specify the physical address and size of the memory
ranges"\cite{DTS} . Furthermore, it is stated, that the property name reg has
the value encoded as a number of (address, length) pairs. It also states, that
the number of <u32> cells required to specify the address and length are
bus-specific and are specified by the \#address-cells and \#size-cells properties
in the parent of the device node.
Looking through our riscv32.dts file, we find the relevant information to be:
\begin{lstlisting}
#address-cells = <0x02>;
#size-cells = <0x02>;

memory@80000000 {
  device_type = "memory";
  reg = <0x00 0x80000000 0x00 0x8000000>
};
\end{lstlisting}
With the information previously provided, we know that the starting
address of the memory section is at address $0x00 + 0x80000000 = 0x80000000$ and has a size of
$0x00 + 0x8000000$ bytes, which is equivalent to 128MB.

\subsubsection{Creating a linker script}
The linker script is used to tell the linker which parts of the file to include
in the final output file, aswell as where each section is stored in memory. As
we are working on an embedded system, we have to stray from the default and
create our own linker script. The clang uses the llvm lld linker, which is compatible
with the general linker scripts implementations of the GNU ld linker\cite{llvm-org-linker}.
Thus, we can make use of the GNU ld manual for modifying the linker script in freeRTOS for our bare metal application instead of writing the entire thing from scratch\cite{GNU-linker}.

\begin{lstlisting}
OUTPUT_ARCH('riscv')
ENTRY(_start)

MEMORY
{
/* Fake ROM area */
rom (rxa) : ORIGIN = 0x80000000, LENGTH = 1M
ram (wxa) : ORIGIN = 0x80100000, LENGTH = 127M
}
\end{lstlisting}
First, we must specify that we want the RISC-V architecture and designate the entry point
of the program at a function named "\_start," which we will define later.
Second, we define the MEMORY area to consist of both a writable memory region and a read-only
memory region. We name these regions "ram" and "rom," respectively. With that we move on to define the SECTIONS element of the linker script.

\begin{lstlisting}
SECTIONS
{
.text : ALIGN(CONSTANT(MAXPAGESIZE))
{
  *(.text .text.*)
}

.rodata : ALIGN(CONSTANT(MAXPAGESIZE))
{
  *(.rdata)
  *(.rodata .rodata.*)
}

.data : ALIGN(CONSTANT(MAXPAGESIZE))
{
  *(.data .data.*)
  /*RISCV convention to have __global_pointer
  aligned to 8 bytes*/
  . = ALIGN(8);
  PROVIDE( __global_pointer$ = . + 0x800 );
}

.bss : ALIGN(CONSTANT(MAXPAGESIZE))
{
  *(.bss .bss.*)
}

/* It is standard to have
the stack aligned to 16 bytes*/
. = ALIGN(16);
_end = .;

.stack : ALIGN(CONSTANT(MAXPAGESIZE))
{
  _stack_top = ORIGIN(ram) + LENGTH(ram);
}
}
\end{lstlisting}
The text, rodata, data and bss sections follows the same general procedure. We align the
section to the maximum size of a page, and the match all the data which we care about for
the given sections. I have opted to disregard specifying where the linker has to save
all the data, and instead opted to let the linker itself find a suited location looking at
the attributes we gave to memory previously.
In the data section, we also provide a global pointer, which is used to access global variables
within our later code implementation.\footnote{In general the global pointer is used together
with an offset in much the same manner as a stack pointer and offset.}

Then I align the end with 16 bytes as is the custom. Reason this is moved outside the stack
is that I use the \_end variable later, and as such it should be aligned aswell.
Then within the stack section we define the \_stack\_top as being the end of the random access
memory section (ram), as the stack grows downwards.

\subsubsection{Getting into the main function}
In the linker script we specified the entry point of our program as \_start. Next up
is implementing said entry point in assembly. Within a new assembly file we add the following.
\begin{lstlisting}
.extern main
.section .init
.globl _start
.type _start,@function

_start:
  .cfi_startproc
  .cfi_undefined ra
  .option push
  .option norelax
  la gp, __global_pointer$
  .option pop
  // load _stack_top into the sp register
  la sp, _stack_top
  add s0, sp, zero

  // argc, argv, envp is 0 and jump to main
  li  a0, 0
  li  a1, 0
  li  a2, 0
  jal main
  .cfi_endproc // We end the proccess
\end{lstlisting}
First we specify that externaly there will be implemented a main entry point. Next we tell the
linker to save the following code in the .init section and initialize a global label \_start,
and note that it is a function.

Next the \_start is defined, and define .cfi\_startproc such that we have an entry in the
.eh\_frame. Next we define the return address register(ra) as being undefined, as we are in
the start of the entire program. Since the linker usually relaxes adressing sequences to
shorter GP-relative sequences when possible, the initial load of GP must not be relaxed
\cite{GNU_BIN}. However, we do not need the same for loading the \_stack\_top into the sp
register, and then also save it in the s0 register, which stores the frame pointer.
Then the last step is loading 0 into argc, argv and envp and the jump to the externaly
defined main function.

\subsubsection{Printing with UART}
\begin{lstlisting}[float=*, label=lst:putchar, caption=Implementation of putchar of stdarg lib]
#include <stdint.h>
#define UART_ADDR 0x10000000
#define LCR 0x03          // Line control register
#define LSR 0x05          // Line status register
#define FCR 0x02          // FIFO control reigster
#define RBR 0x00          // Receiver buffer register
#define IER 0x01          // Interrupt enable register
#define LSR_THRE 0b110000 //
void uart_init(void) {
  volatile uint8_t *ptr = (uint8_t *)UART_ADDR;

  // Set word length to 8 (LCR[1:0])
  *(ptr + LCR) = 0b11;

  // Enable FIFO (FCR[0])
  *(ptr + FCR) = 0b1;

  // Enable receiver buffer interrupts (IER[0])
  *(ptr + IER) = 0b1;
}

static void uart_put(uint8_t c) { *(uint8_t *)(UART_ADDR + RBR) = c; }
static uint8_t uart_get(uintptr_t addr) { return *(uint8_t *)(addr); }

void putchar(unsigned char c) {
  volatile uintptr_t ptr = (uintptr_t)UART_ADDR;
  // make sure there is nothing else in FIFO
  while ((uart_get(ptr + LSR) & LSR_THRE) == 0) {
    // do nothing
  }
  // add the char to reciver buffer register
  uart_put(c);
}
\end{lstlisting}
As we are working with bare metal, the standard printf function will not work. However, QEMU
allows the use of universal asynchronous receiver / transmitter (UART) protocol, which allows
us to implement a printf function, that writes to the terminal without the need for the QEMU
GUI interface. Furthermore, there exists opensource baremetal versions of the printf function,
which makes the effort of printing a lot easier. One such example is Georges Menie's stdarg,
which depends on a single function called putchar, which has to take a character and place it
somewhere. One such implementation can be seen in Listing~\ref{lst:putchar}. Again this code is
a modified version of what is seen in freeRTOS. From the previous dtc file created
in Section~\ref{sec:sys-info} there is also information regarding the UART configuration.
\begin{lstlisting}
serial@10000000 {
  interrupts = <0x0a>;
  interrupt-parent = <0x03>;
  clock-frequency = "\08@";
  reg = <0x00 0x10000000 0x00 0x100>;
  compatible = "ns16550a";
};
\end{lstlisting}
First it states the address, 0x10000000, and it states that it is ns16550a compatible
\cite{uart}. The specifications states the adresses of the different registers seen above.
This allows us to first initialize the uart, and then afterwards define a uart\_put function,
which takes and unsigned character and sends it to the receive buffer register. The putchar
function also makes sure that we do not overwrite a previous character by busy looping until
the FIFO is clear.



