
\section{Related work}
Marcelino et al. \cite{sorting_units} evaluate three hardware sorting units
implemented with specific Field Programmable Gate Arrays (FPGAs). One of these
being the FIFO\footnote{Short for First In First Out}-based merge sorting
machine, where they present a merge sorting structure for a FIFO embedded system
merging. This implementation assumes two sorted input lists and they later
propose a hybrid solution using Insertion and FIFO based merge sorting. With
this they find that the FPGA hybrid insertion and FIFO-based merge sorting sees
speed-ups between 1.6 and 15 time compared to a quick sort pure software
solution. Such a hybrid solution could be implemented in tandem with the
implementation provided in this thesis, and might lead to a similar performance
increase from what they show within their paper.

Jackson et al. \cite{flash_sorting} look for faster sorting algorithms used for
flash memory embedded devices. They provide a merge sort for sorting with
minimal memory usage which aims to reduce the number of WRITES to flash memory.
This implementation would only need two memory buffers. They find that when
sorting large data sets with small memory the proposed algorithms reduces I/Os
and execution time by about 30\%. A design choice in CSD might be to reduce the
need for large amount of memory, and as such a memory efficient sorting
algorithm could be better suited.

Lobo et al. \cite{merge_sort} compare the performance of different types of
merge sort algorithms. Within their testing they find that the serial and
parallel merge sort discussed have a similar amount of resource
utilization, but find the delay of the parallel merge sort to be a lot smaller
than the serial counterpart. As such, they theorize that this implies the
parallel execution to be much faster than the serial. This finding supports the
theory that a parallelized merge sort algorithm, such as the one provided in
this thesis, could increase the performance over a serial counterpart.
